---
title: "Exercise Quality Prediction"
author: "Data Science Learner"
date: "October 25, 2015"
output: html_document

---
### Overview

This analysis is to predict the exercise quality of a group of 6 people, based on the data recorded by the accelerometers tied to their body parts. In this report, we will document the entire analysis, including how to load and preprocess data, how to build a machine learning model, how to implement cross validation, how to estimate the expected out of sample error. It will also explain why such choices were made.  

---
### Data Loading and Preprocessing

First let us load and read the data.
```{r}
if (!file.exists("pml-training.csv")) {
    train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
    download.file(train_url, "pml-training.csv")  
}

if (!file.exists("pml-testing.csv")) {
    test_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
    download.file(test_url, "pml-testing.csv")
}

train_data <- read.csv("pml-training.csv")
```

Then let us explore the data.
```{r}
dim(train_data)
table(train_data$classe)
```

We can also examine the summary of the training data set by *summary(train_data)*, but due to the large number of variables, we won't display its result here. This summary provides very valuable information. Some columns are meaningless, and some have an enormous number of missing values. Therefore, I collected these columns and decided to remove them from analysis.

```{r}
cols_rm <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2",
             "cvtd_timestamp", "new_window", "num_window",
             "kurtosis_roll_belt", "kurtosis_picth_belt", 
             "kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", 
             "skewness_yaw_belt", "max_roll_belt", "max_picth_belt", 
             "max_yaw_belt", "min_roll_belt", "min_pitch_belt", "min_yaw_belt", 
             "amplitude_roll_belt", "amplitude_pitch_belt",
             "amplitude_yaw_belt", 
             "var_total_accel_belt", "avg_roll_belt", "stddev_roll_belt", 
             "var_roll_belt", "avg_pitch_belt", "stddev_pitch_belt", 
             "var_pitch_belt", "avg_yaw_belt", "stddev_yaw_belt",
             "var_yaw_belt", 
             "var_accel_arm", "avg_roll_arm", "stddev_roll_arm", "var_roll_arm",
             "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm",
             "avg_yaw_arm", 
             "stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", 
             "kurtosis_picth_arm", "kurtosis_yaw_arm", "skewness_roll_arm", 
             "skewness_pitch_arm", "skewness_yaw_arm", "max_roll_arm", 
             "max_picth_arm", "max_yaw_arm", "min_roll_arm", "min_pitch_arm", 
             "min_yaw_arm", "amplitude_roll_arm", "amplitude_pitch_arm", 
             "amplitude_yaw_arm", "kurtosis_roll_dumbbell", 
             "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", 
             "skewness_roll_dumbbell", "skewness_pitch_dumbbell", 
             "skewness_yaw_dumbbell", "max_roll_dumbbell", "max_picth_dumbbell",
             "max_yaw_dumbbell", "min_roll_dumbbell", "min_pitch_dumbbell",
             "min_yaw_dumbbell", "amplitude_roll_dumbbell",
             "amplitude_pitch_dumbbell", 
             "amplitude_yaw_dumbbell", "var_accel_dumbbell",
             "avg_roll_dumbbell", 
             "stddev_roll_dumbbell", "var_roll_dumbbell", "avg_pitch_dumbbell", 
             "stddev_pitch_dumbbell", "var_pitch_dumbbell", "avg_yaw_dumbbell", 
             "stddev_yaw_dumbbell", "var_yaw_dumbbell", "kurtosis_roll_forearm",
             "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
             "skewness_roll_forearm", 
             "skewness_pitch_forearm", "skewness_yaw_forearm",
             "max_roll_forearm", 
             "max_picth_forearm", "max_yaw_forearm", "min_roll_forearm", 
             "min_pitch_forearm", "min_yaw_forearm", "amplitude_roll_forearm", 
             "amplitude_pitch_forearm", "amplitude_yaw_forearm",
             "var_accel_forearm", 
             "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm", 
             "avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", 
             "avg_yaw_forearm", "stddev_yaw_forearm", "var_yaw_forearm"
             )

train_data1 <- train_data[, !names(train_data) %in% cols_rm] 
dim(train_data1)
```


### Building a Model 

The random forest method is said to be a good tool for prediction, therefore I applied it in this analysis. For the prediction accuracy concern, we need to cross validate the model, and I chose to use the 10-fold cross validation method. Below is the code to build the model and implement cross validation.

```{r, cache = TRUE}
library(caret)
library(randomForest)

set.seed(52890)
knum <- 10 ## k folds
folds <- createFolds(train_data1$classe, k = knum, list = TRUE, returnTrain = FALSE)
sapply(folds, length)

rows <- 1:dim(train_data1)[1]
fnl_col <- dim(train_data1)[2]  ## the final column number

err_rate_oob <- rep(NA, knum)  ## record the estimated error rate from training
err_rate_test <- rep(NA, knum) ## record the estimated error rate from test

tnum <- 300 ## number of trees to build in each iteration

for (i in 1:knum) {
    training <- train_data1[!(rows %in% folds[[i]]),]
    testing <- train_data1[rows %in% folds[[i]],]
    rf <- randomForest(classe ~ ., data = training, prox = FALSE, ntree = tnum,
                       xtest = testing[, -fnl_col], ytest = testing$classe)
    err_rate_oob[i] <- rf$err.rate[tnum, "OOB"]
    kk <- rf$test
    err_rate_test[i] <- kk$err.rate[tnum, "Test"]
}
```

We can obtain the two average estimated error rates for predicting out-of-bag data and test data, respectively. They are as follows.
```{r}
mean(err_rate_oob)
mean(err_rate_test)
```

Therefore, we can say, we expect the out of sample error rate to be `r mean(err_rate_oob)`, and the estimated error rate from cross validation is `r mean(err_rate_test)`.


### Results

Both the two error rates are very small, so the prediction model using the random forest method is excellent.



